Building and Checking Suffix Array Using Induced Sorting Method

### 2017-04-25 new-draft/draft.tex
2017/04/25

In the past decade, the induced sorting~(IS) method has been successively used to design disk-based suffix array~(SA) construction algorithms that have better time and I/O complexities than the existing alternatives. However, their current programs commonly suffer from a performance bottleneck due to large disk space in need. Recently, it was reported a carefully engineered IS suffix sorter that takes less than $8n$ disk space for constructing an SA encoded by 40-bit integers. This indicates a great potential for improving space efficiency of the prior arts, such as DSA-IS and SAIS-PQ. In this paper, we present two checking methods that enable any IS suffix sorting algorithms to check an SA when it is being built, where the second method can also be applied to ensure the correctness of an LCP array. For performance analysis, we integrate the proposed methods into DSA-IS and implement the enhanced algorithm to evaluate the checking overhead, where the program design of DSA-IS has been optimized for a better speed and smaller peak disk use. From our experiments, the time, space and I/O consumptions for verification is negligible in comparison with that for construction. In view of the high performance of these checkers and the existing IS suffix sorters, we are convinced that our checking methods, in accompanied with the state-of-the-art IS suffix sorting algorithm, constitute efficient solutions to the situations where checking is a must after construction.

@@

Given a string drawn from a constant or integer alphabet, its suffix array can be built by the internal-memory algorithm SA-IS~\cite{Nong11} in linear time and space. So far, the IS method has been also applied to designing three external-memory suffix sorting algorithms~eSAIS~\cite{Bingmann12}, DSA-IS~\cite{Nong15} and SAIS-PQ~\cite{Liu15}, which have better time and I/O complexities than other alternatives, e.g., DC3~\cite{Dementiev2008a}, bwt-disk~\cite{Ferragina2012}, SAscan~\cite{Karkkainen2014} and pSAscan~\cite{Karkkainen2015}. However, the peak disk uses of their current programs are at least 15 times the sizes of input strings, resulting in a performance bottleneck. Recently, it was reported in~\cite{Karkkainen2017} a new IS suffix sorting algorithm fSAIS with a careful engineering that consumes nearly optimal disk space and runs faster than the parallel suffix sorter pSAscan on massive datasets. The algorithmic design of fSAIS is highly dependent on that of the prior arts. Particularly, it employs a sorting method similar to that of DSA-IS for inducing suffixes/substrings and the naming method of SAIS-PQ for producing reduced strings. The substantial improvement on space efficiency is mainly achieved by means of a monotone priority queue  for sorting substrings/suffixes and a garbage collector for quickly recycling disk space occupied by data no longer to use. As a result, its implementation takes less than $8n$ disk space for any input string of a size $n \le 2^{40}$, indicating a great potential for optimizing the programs for DSA-IS and SAIS-PQ. In this paper, we make our first attempt to redesign DSA-IS using new substring sorting and naming methods. The experimental results indicate that our program for the enhanced algorithm consumes about two-thirds as much disk space as that for eSAIS and runs faster than the latter on real world datasets. In addition, we also give a discussion on further improving our program to achieve a better space efficiency using the techniques for implementing fSAIS.

To detect potential computation errors caused by implementation bugs and/or hardware malfunctions, a constructed SA should be verified to ensure its correctness before use. This checking function has been included into various widespread software packages like SA-IS, DC3 and eSAIS. Particularly, the latter two verify an SA using the checking method presented in~\cite{Dementiev2008a}, which mainly consists of two passes of external-memory sorts involving $\mathcal{O}(n)$ fixed-size tuples with an integer key. In this paper, we propose a new checking method that enables any IS suffix sorting algorithm to check an SA when it is being built. 
For performance analysis, we integrate the proposed method into the optimized DSA-IS and implement the combined algorithm, called DSA-IS+, to evaluate the checking overhead. From our experiments, the time, space and I/O consumptions for verification is negligible in comparison with that for construction. In view of the high performance of this checker and the existing IS suffix sorters, we believe that our checking method, together with the state-of-the-art IS suffix sorting algorithm, can constitute an efficient solution to the situations where checking is a must for a constructed SA.

The rest of the paper is organized as follows. We first describe in Sections~\ref{sec:build_sa} and~\ref{sec:check_sa} the adapted algorithmic design of DSA-IS and the proposed checking method, respectively. Then, we demonstrate the experimental results in Section~\ref{sec:experiments} and draw the concluding remarks in Section~\ref{sec:conclusion}.

>>>>

In the past decade, the induced sorting~(IS) method has been successively used to design disk-based suffix array~(SA) construction algorithms that have better time and I/O complexities than the existing alternatives. However, their current programs commonly suffer from a performance bottleneck due to large disk space in need. Recently, it was reported a carefully engineered IS suffix sorter that takes less than $8n$ disk space for constructing an SA encoded by 40-bit integers. This indicates a great potential for improving space efficiency of the prior arts, such as DSA-IS and SAIS-PQ. We present two checking methods that enable any IS suffix sorting algorithms to check an SA when it is being built, where the second method can also be applied to ensuring the correctness of an LCP array. For performance analysis, we integrate the proposed methods into DSA-IS and implement the enhanced algorithm to evaluate the checking overhead, where the program design of DSA-IS has been optimized in advance for a better speed and smaller peak disk use. From our experiments, the time, space and I/O consumptions for verification is negligible in comparison with that for construction. In view of the high performance of these checkers and the existing IS suffix sorters, we are convinced that our checking methods, in accompanied with the state-of-the-art IS suffix sorting algorithm, constitute efficient solutions for the situations where checking is a must after construction.

@@

Given a string drawn from any constant or integer alphabet	, its suffix array can be constructed in linear time and space by the internal-memory algorithm SA-IS~\cite{Nong11}. So far, the IS method has been also applied to designing three external-memory suffix sorting algorithms~eSAIS~\cite{Bingmann12}, DSA-IS~\cite{Nong15} and SAIS-PQ~\cite{Liu15}. Theoretically, these algorithms have better time and I/O complexities than the existing alternatives, e.g., DC3~\cite{Dementiev2008a}, bwt-disk~\cite{Ferragina2012}, SAscan~\cite{Karkkainen2014} and pSAscan~\cite{Karkkainen2015}, but they all suffer from a space bottleneck due to the large disk space in use. Recently, it was reported in~\cite{Karkkainen2017} a new IS suffix sorting algorithm fSAIS with a careful engineering that achieves optimal peak disk use and fastest speed among the state-of-the-art suffix sorters on massive datasets. The algorithm design of fSAIS is highly dependent on that of the prior arts, where it uses a sorting method similar to that of DSA-IS for inducing suffixes/substrings and the naming method of SAIS-PQ for producing reduced strings. Its high space performance is mainly due to the use of a monotone priority queue for sorting substrings/suffixes and a garbage collector for quickly recycling disk space occupied by temporary data no longer to use. As a result, its implementation takes less than $8n$ disk space for any input string of a size $n \le 2^{40}$, which also indicates a great potential for optimizing the program designs of DSA-IS and SAIS-PQ. In this paper, we present our first attempt to redesign DSA-IS by means of new substring sorting and naming methods. The experimental results indicate that our program for the enhanced algorithm consumes about two-thirds as much disk space as that for eSAIS and runs faster than the latter on real world datasets. The enhanced program can be further improved to achieve better time and space performance using the techniques for implementing fSAIS.

To detect potential computation errors caused by implementation bugs and/or hardware malfunctions, a suffix array should be verified before use. This checking function has been included into various widespread software packages like SA-IS, DC3 and eSAIS. Particularly, the latter two verify an SA using the checking method presented in~\cite{Dementiev2008a}, which mainly consists of two passes of external-memory sorts involving $\mathcal{O}(n)$ fixed-size tuples with an integer key. In this paper, we propose two checking methods that enable any IS suffix sorting algorithm to verify an SA when it is being built, where the second method can also be applied to ensuring the correctness of an LCP array. For performance analysis, we integrate the proposed checking method into the optimized DSA-IS and implement the combined algorithms to evaluate the checking overhead. From our experiments, the time, space and I/O consumptions for verification is negligible in comparison with that for construction. In view of the high performance of these checkers and the existing IS suffix sorters, we believe that our checking methods, together with the state-of-the-art IS suffix sorting algorithm, can constitute efficient solutions for the situations where building and checking must be performed at the same time.

The rest of this paper is organized as follows. We first describe in Sections~\ref{sec:build_sa} and~\ref{sec:check_sa} the adapted algorithmic design of DSA-IS and the proposed checking methods, respectively. Then, we demonstrate the experimental results in Section~\ref{sec:experiments} and draw the conclusion in Section~\ref{sec:conclusion}.
